Place of death in NRS COVID-19 deaths

### 2021-02-03 covid_nrs_place_of_death.Rmd
added crown copyright notice; modifying age/sex data split in progress

@@

~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above!

@@

weekly_deaths_age_sex <-
  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021, we only want the weekly figures
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  select(-c(measure_type)) %>%  #  measure_type is count for all data
  recode_place_of_death() %>%
  recode_cause_of_death() %>%
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing

@@

```{r}
weekly_deaths_age_sex %>% 
  count(ref_period, date_w_c, week_number, week_number_run_over) %>% 
  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
  select(-n) %>% 
  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
```

@@

past_deaths_age_sex_file <-
  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
                destfile = tempfile(),
                quiet = FALSE)

@@

past_deaths_hb <- 
  readxl::read_excel(past_deaths_health_board_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(
    week_number = week_of_occurrence,
    place_of_death = location
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    health_board, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

past_deaths_la <-
  readxl::read_excel(past_deaths_council_area_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(
    week_number = week,
    place_of_death = location_of_death
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    council_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

past_deaths_age_sex <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(past_deaths_age_sex_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  mutate(
    year = 2000L + as.integer(year),
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()

## check that the results are the same as overall deaths
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_health_board, past_deaths_council_area, past_deaths_age_sex),
  .y = c("overall","health_board","council_area","age_sex"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(!is.na(year) & place_of_death!="All locations") %>%
      group_by(dataset,year) %>%
        summarise(total = sum(number_of_deaths), .groups = "drop")
  }
) %>% 
  pivot_wider(names_from=year,values_from=total) %>%
  knitr::kable(caption = "Number of deaths by year as calculated from different datasets. All datasets apart from the split by health board agree. The differences in the health board dataset likely reflect missing values and/or small numbers redacted from the dataset, which may have to do with the grouping of state hospital/non-NHS institutions.")
```

@@

```{r, warning=FALSE}
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_la, past_deaths_age_sex),
  .y = c("overall","health_board","council_area","age_sex"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(year %in% 2016:2019 & week_number==53) %>%
      count(dataset,year)
  }
) %>% 
  pivot_wider(names_from=year,values_from=n) %>%
  knitr::kable(caption = "Number of entries by dataset for week number 53 in years 2016-2019 (which only had 52 weeks). Only the 'overall' historical deaths dataset has week 53 entries for years with 52 weeks, and all the entries are for 0 deaths.")

@@

# TODO: clean up once these have been dealth with

@@

historical_average_age_sex <-
  past_deaths_age_sex %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, age, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_age <-
  past_deaths_age_sex %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, age
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_sex <-
  past_deaths_age_sex %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
```

@@

```{r}
weekly_deaths_age_sex %>%
  filter(sex=="all" & age=="all" & cause_of_death=="all-causes-average-of-corresponding-week-over-previous-5-years") %>%
  select(week_number, year = year, place_of_death, nrs_computed_mean = number_of_deaths) %>%
  left_join(historical_average_overall) %>%
  filter(round(nrs_computed_mean) != round(deaths_mean)) %>%
  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file. Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019. For all the other datasets the mean deaths computed agree!")
```

@@

```{r}
merged_data_overall <-
  weekly_deaths_age_sex %>%
  filter(sex=="all" & age=="all") %>%
  # covid_deaths$scotland %>%
  split_deaths_due_to_covid_and_not %>%
  left_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death", "week_number", "sex", "age"))

@@

# merged_deaths_by_age_sex <-
#   deaths_by_cause %>%
#   split_deaths_due_to_covid_and_not %>%
#   left_join(historical_range_by_age_sex) %>%
#   compute_run_over_week_for_2021
# 
# merged_deaths_by_age <-
#   deaths_by_cause %>%
#   split_deaths_due_to_covid_and_not %>%
#   left_join(historical_range_by_age) %>%
#   compute_run_over_week_for_2021
# 
# merged_deaths_by_sex <-
#   deaths_by_cause %>%
#   split_deaths_due_to_covid_and_not %>%
#   left_join(historical_range_by_sex) %>%
#   compute_run_over_week_for_2021
```

>>>>

# Copyright attribution

All data used for this study were obtained from the National Records Scotland (NRS) and are © Crown copyright, 2020; the details of the licence can be viewed on the [Open Government Licence website](http://www.nationalarchives.gov.uk/doc/open-government-licence/open-government-licence.htm)

@@

~~Using `opendatascot` to load weekly deaths the level of geography can be specified, so we can have more or less granular data~~ Choosing a more granular level than Scotland-wide data means place of death is collapsed to the 'all' category only, see note above! Even for the Scotland-wide data, place of death is collapsed to 'All' for the granular sex & age categories, actually.

@@

weekly_deaths_overall <-
  ods_dataset("deaths-involving-coronavirus-covid-19", geography = "sc") %>%
  make_human_readable_ref_area %>%
  clean_names() %>%
  ## below series of rename operations are a stupid way of dealing with the possibility of different variable name conventions in the ods_dataset data: sometimes they are in snake_case, and sometimes in camelCase, seemingly unpredictably
  rename_at(.vars = vars(matches("location", ignore.case = TRUE)), ~"place_of_death") %>%
  rename_at(.vars = vars(matches("period", ignore.case = TRUE)), ~"ref_period") %>%
  rename_at(.vars = vars(matches("area", ignore.case = TRUE)), ~"ref_area") %>%
  rename_at(.vars = vars(matches("measure", ignore.case = TRUE)), ~"measure_type") %>%
  rename_at(.vars = vars(matches("cause", ignore.case = TRUE)), ~"cause_of_death") %>%
  rename_at(.vars = vars(matches("sex", ignore.case = TRUE)), ~"sex") %>%
  rename_at(.vars = vars(matches("age", ignore.case = TRUE)), ~"age") %>%
  rename_at(.vars = vars(matches("value", ignore.case = TRUE)), ~"number_of_deaths") %>%  # rename so it's consistent with other data
  filter(!ref_period %in% as.character(2020:2021)) %>% # remove totals for 2020 & 2021, we only want the weekly figures
  mutate(
    year = parse_integer(str_extract(ref_period, pattern="^(2020|2021)")),
    week_number = parse_integer(str_remove(ref_period, pattern = "^(2020|2021)\\-")),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    number_of_deaths = as.numeric(number_of_deaths),  # convert to numeric!
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  select(-c(measure_type)) %>%  #  measure_type is count for all data
  recode_place_of_death() %>%
  recode_cause_of_death() %>%
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%  # this dataset is well-formatted and has no implicit missing values!
  mutate(deaths_non_covid = deaths_all_causes - deaths_covid_related)  # compute non-covid deaths from the other two figures - will be NA where either is missing

weekly_deaths_sex_age_file <-
  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-sex.xlsx",
                destfile = tempfile(),
                quiet = FALSE)

weekly_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(weekly_deaths_sex_age_file, sheet = 2, skip = 4, col_names = c("week_number","place_of_death","sex","age","cause_of_death","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  mutate(
    year = 2000L + parse_integer(str_sub(week_number, 1, 2)),
    week_number = parse_integer(str_sub(week_number, 4, 5)),
    date_w_c = compute_start_date_from_week_number(week_number = week_number, year = year),
    week_number_run_over = if_else(year == 2021, 53L + week_number, week_number)  # invent a "run-over" week number - weeks 54+ are in 2021
  ) %>%
  recode_cause_of_death() %>%
  tidyr::complete(
    nesting(week_number,week_number_run_over,date_w_c,year),
    age,sex,place_of_death,cause_of_death,
    fill = list(number_of_deaths = 0)
  ) %>%
  recode_place_of_death() %>%  # this needs to be done after recoding cause of death, otherwise we also include the "all" figure which isn't included in this dataset
  pivot_wider(names_from = cause_of_death, values_from = number_of_deaths) %>%   # make separate columns for causes of death figures
  mutate(deaths_all_causes = deaths_non_covid + deaths_covid_related) %>%  # compute all cause deaths from the other two figures reported
  set_covid_and_non_covid_deaths_to_zero_before_reporting_started()

@@

```{r}
weekly_deaths_overall %>% 
  count(ref_period, date_w_c, week_number, week_number_run_over) %>% 
  slice(1:3, (nrow(.) - 3):nrow(.)) %>% 
  select(-n) %>% 
  knitr::kable(caption = "Checking the computation of dates from week numbers - note how the 1st week of 2020 started in 2019.")
```

@@

past_deaths_sex_age_file <-
  curl_download(url = "https://www.nrscotland.gov.uk/files//statistics/covid19/weekly-deaths-by-location-age-group-sex-15-19.xlsx",
                destfile = tempfile(),
                quiet = FALSE)

@@

past_deaths_hb <- 
  readxl::read_excel(past_deaths_hb_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(
    week_number = week_of_occurrence,
    place_of_death = location
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    health_board, place_of_death,  # fill in missing combinations of hb & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

past_deaths_la <-
  readxl::read_excel(past_deaths_la_file, sheet = 2, skip = 2, col_names = TRUE) %>%  # first two rows are empty, row 3 has headers
  clean_names() %>%
  drop_na(number_of_deaths) %>%  # remove empty rows & one row where it's just the copyright notice
  rename(
    week_number = week,
    place_of_death = location_of_death
  ) %>%
  mutate(
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    council_area, place_of_death,  # fill in missing combinations of la & pod with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()  # recode place of death after completing missing combinations, otherwise the factor level "All" gets exploded also

past_deaths_sex_age <-  # structure here is a bit more complicated - headers in row 3, but sex & age headers in row 4
  readxl::read_excel(past_deaths_sex_age_file, sheet = 2, skip = 4, col_names = c("year","week_number","place_of_death","sex","age","number_of_deaths"), guess_max = 1e5) %>%
  drop_na(number_of_deaths) %>%    # remove empty rows & one row where it's just the copyright notice
  mutate(
    year = 2000L + as.integer(year),
    week_number = as.integer(week_number)
  ) %>%
  tidyr::complete(
    nesting(week_number,year),  # unique combinations that we don't want to combinatorily explode
    sex, age, place_of_death,  # fill in missing combinations of age, sex, pod, with 0 deaths
    fill = list(number_of_deaths = 0)
  ) %>% 
  recode_place_of_death()

## check that the results are the same as overall deaths
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_hb, past_deaths_sex_age),
  .y = c("overall","health_board","council_area","sex_age"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(!is.na(year) & place_of_death!="All") %>%
      group_by(dataset,year) %>%
        summarise(total = sum(number_of_deaths), .groups = "drop")
  }
) %>% 
  pivot_wider(names_from=year,values_from=total) %>%
  knitr::kable(caption = "Number of deaths by year as calculated from different datasets. The overall dataset and age/sex breakdown agree in figures, and the HB and LA breakdowns agree, but the two pairs differ from each other. The difference is in both directions, some years there are more deaths recorded in one pair, and some years in the other. Health Board and Local Authority figures include non-residents. Deaths are allocated to areas based on the usual residence of the deceased. If the deceased was not a Scottish resident, the death is allocated to the area where the death occurred.")
```

@@

```{r, warning=FALSE}
map2_dfr(
  .x = list(past_deaths_overall, past_deaths_hb, past_deaths_la, past_deaths_sex_age),
  .y = c("overall","health_board","council_area","sex_age"),
  .f = function(data_tbl, name) {
    data_tbl %>%
      mutate(dataset=name,year=as.numeric(year)) %>%
      filter(year %in% 2016:2019 & week_number==53) %>%
      count(dataset,year)
  }
) %>% 
  pivot_wider(names_from=year,values_from=n) %>%
  knitr::kable(caption = "Number of entries by dataset for week number 53 in years 2016-2019 (which only had 52 weeks). Only the 'overall' historical deaths dataset has week 53 entries for years with 52 weeks, and all the entries are for 0 deaths.")

@@

@@

historical_average_sex_age <-
  past_deaths_sex_age %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, age, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_average_age <-
  past_deaths_sex_age %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, age
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )

historical_aversex_age <-
  past_deaths_sex_age %>%
  create_week_53_data_for_years_with_52_weeks() %>%
  group_by(
    week_number, place_of_death, sex
  ) %>%
  summarise(
    deaths_mean = mean(number_of_deaths),
    deaths_min = min(number_of_deaths),
    deaths_max = max(number_of_deaths),
    .groups = "drop"
  )
```

@@

```{r}
weekly_deaths_overall %>%
  filter(sex=="all" & age=="all") %>%
  left_join(historical_average_overall) %>%
  select(year, week_number, place_of_death, deaths_nrs_past_average_all_causes, deaths_mean) %>%
  filter(round(deaths_nrs_past_average_all_causes) != round(deaths_mean)) %>%
  knitr::kable(caption = "Subset of weekly deaths where calculated mean deaths 2015-2019 differed between the COVID-19 related deaths file and the weekly deaths file (i.e. all other entries had the same figures!). Week 53 was computed as the average of deaths in week 53 in 2015 and week 52 in 2016-2019. For all the other datasets the mean deaths computed agree!")
```

@@

```{r}
merged_data_overall <-
  weekly_deaths_sex_age %>%
  filter(sex=="all" & age=="all") %>%
  inner_join(historical_average_overall %>% mutate(sex = "all", age = "all"),  # only merge for aggregated data b/c that's all we have!
            by = c("place_of_death", "week_number", "sex", "age"))

@@

merged_data_sex_age <-
  weekly_deaths_sex_age %>%
  inner_join(historical_average_sex_age, by = c("place_of_death", "week_number", "sex", "age"))

# merged_deaths_by_sex_age <-
#   deaths_by_cause %>%
#   split_deaths_due_to_covid_and_not %>%
#   left_join(historical_range_by_sex_age) %>%
#   compute_run_over_week_for_2021
# 
# merged_deaths_by_age <-
#   deaths_by_cause %>%
#   split_deaths_due_to_covid_and_not %>%
#   left_join(historical_range_by_age) %>%
#   compute_run_over_week_for_2021
# 
# merged_deaths_by_sex <-
#   deaths_by_cause %>%
#   split_deaths_due_to_covid_and_not %>%
#   left_join(historical_range_by_sex) %>%
#   compute_run_over_week_for_2021
```
